Multi-scale trend analysis of water quality using error propagation of generalized additive models

### 2020-11-09 manu_draft.Rmd
results text updated

Examples of results provided by the trend tests are shown for station 34, with estimates of percent change between year pairs and seasonal trend results for January to June and July to December for the same year pairs (Figure \@ref(fig:trnddat)).  This figure illustrates the narrative descriptions that can be obtained for each test and the differences that can be observed depending on season.  Figures \@ref(fig:trnddat)a-c demonstrate that average chlorophyll at station 34 from 1990 to 2000 had a 268,4% increase, although the change is insignificant. Similarly, the change from 2000 to 2010 was an insignificant increase of 11%, followed by an insignificant decrease of 13.9% from 2010 to 2019.  Although these trends were insignificant, the results are based on a priori decisions of which years to compare.  Significant results may be observed for other year pairs.

@@

Spatial variation in trends suggested that most stations in South Bay had increasing chlorophyll from 1990 to 2010 following a decrease in recent years, although many trends were not significant and varied by location (Figure \@ref(fig:trndmap)a).  From 1990 to 2000, only station 18 and 22 in the northern part of South Bay had significant increasing trends.  The magnitude of the chlorophyll increase in the southernmost stations (34, 36) was higher during this same period, although the trends were not significant.  From 2000 to 2010, significant increases were observed at the three northern-most stations (18, 21, and 22) and two southern stations (32, 34).  From 2010 to 2016, all four southern stations had significant decreases in annual chlorophyll.  The spatial trends were further evaluated with mixed-meta regression analysis to provide additional context to the annual changes by evaluating trends from January to July and August to December (Figure \@ref(fig:trndmap)b).  For the January to July period, significant changes were only observed at the southern stations, with increases at station 34 and 36 from 1990-2000, an increase at station 34 from 2000-2010, and a decrease at station 32 form 2010 to 2016.  No significant changes were observed for the August to December period.

A final analysis of the seasonal trends evaluated more discrete monthly periods in three month blocks to provide finer resolution to explain trends from the GAM results (Table \@ref(tab:seastrndtab)).  Many stations and time periods did not show any seasonal chlorophyll trends or only had one significant trend (e.g., stations 21 and 24, JFM period from 2000-2010).  However, several trends were observed in the southern stations, with two significant trends at station 32, six at station 34, and three at station 36.  Trends at station 32 indicated a positive increase in the JFM (January, February, March) period from 2000-2010, followed by a significant decrease for the same seasonal period from 2010-2016.  Station 34 had significant increases in the JFM period from 1990-2010 and all other periods the OND (October, November, December) period for 2000 to 2010.  Significant decreases were observed at station 34 in the JFM and AMJ (April, March, July) period from 2010 to 2016.  Station 36 had significant chlorophyll increases in all seasons, except OND for the 1990 to 2000 period.  Accordingly, table \@ref(tab:seastrndtab) demonstrates the value of obtaining greater insight into seasonal changes by evaluating different blocks as compared to the larger seasonal blocks shown in Figure \@ref(fig:trndmap)b.

@@

Trends results from the mixed-meta regression method for each season and different time periods were compared to alternative trend analyses to demonstrate different conclusions that can be derived depending on method.  Figure \@ref(fig:trndcmp) shows estimates of chlorophyll change per year for the same seasonal periods and annual groupings as in figure \@ref(fig:trndmap)b.  In addition to mixed-meta analysis, two alternative trend methods were used where the first shows slope estimates from a linear regression on annual averages of observed chlorophyll for the specified seasonal period and the second shows slope estimates from a linear regression model on annual averages derived from the GAM seasonal estimates.  The latter uses the same averages from the mixed-meta analysis, but is a simpler approach that does not account for the uncertainty in the average estimates.

The different trend analysis methods provided conflicting information on the magnitude and direction of the seasonal chlorophyll changes in each decade (Figure \@ref(fig:trndcmp)).  The slope estimates from the linear model applied to the observed data were understandably more variable than the slope estimate from the average and mixed-meta methods, with much larger slopes observed especially at the more southern stations.  Slope estimates from the linear model applied to the averages from the GAMs as compared ot the mixed-meta results were identical at almost all stations, excluding station 32 in the January-July period from 1990-2000 when the former showed a negative slope and the latter showed a positive slope.  The mixed-meta analyses had less significant trends compared to the average results for the same GAM estimates, which reflects the ability of the former to account for uncertainty in the average estimates when considering trends.  Importantly, completely opposite and/or varying significance of trends were observed between the methods, suggesting choice of analysis method can influence the certainty of conclusions.

@@

@@

p1 <- show_perchg(mods = toplo, baseyr = 1991, testyr = 2000, ylab = ylab) +
  labs(
    subtitle = paste0('(a) 1991: ', chg[[1, 2]],' ug/L, 2000: ', chg[[1, 3]], 'ug/L\nChange: ', chg[[1,4]], '%, ', chg[[1,5]]), 
    title = NULL
  )
p2 <- show_perchg(mods = toplo, baseyr = 2000, testyr = 2010, ylab = NULL) +
  labs(
    subtitle = paste0('(a) 2001: ', chg[[2, 2]],' ug/L, 2010: ', chg[[2, 3]], 'ug/L\nChange: ', chg[[2, 4]], '%, ', chg[[2, 5]]), 
    title = NULL
  )
p3 <- show_perchg(mods = toplo, baseyr = 2010, testyr = 2019, ylab = NULL) +
  labs(
    subtitle = paste0('(a) 2011: ', chg[[3, 2]],' ug/L, 2019: ', chg[[3, 3]], 'ug/L\nChange: ', chg[[3, 4]], '%, ', chg[[3, 5]]), 
    title = NULL
  )
p4 <- show_avgseason(mods = toplo, doystr = 41, doyend = 213, yrstr = 1991, yrend = 2000, ylab = ylab) +
  coord_cartesian(ylim = ylim) + 
  labs(
    subtitle = seas[[1]],
    title = NULL
  )
p5 <- show_avgseason(mods = toplo, doystr = 41, doyend = 213, yrstr = 2000, yrend = 2010, ylab = NULL) +
  coord_cartesian(ylim = ylim) + 
  labs(
    subtitle = seas[[2]], 
    title = NULL
  )
p6 <- show_avgseason(mods = toplo, doystr = 41, doyend = 213, yrstr = 2010, yrend = 2019, ylab = NULL) +
  coord_cartesian(ylim = ylim) +  
  labs(
    subtitle = seas[[3]], 
    title = NULL
  )
p7 <- show_avgseason(mods = toplo, doystr = 213, doyend = 338, yrstr = 1991, yrend = 2000, ylab = ylab) +
  coord_cartesian(ylim = ylim) + 
  labs(
    subtitle = seas[[4]], 
    title = NULL
  )
p8 <- show_avgseason(mods = toplo, doystr = 213, doyend = 338, yrstr = 2000, yrend = 2010, ylab = NULL) +
  coord_cartesian(ylim = ylim) + 
  labs(
    subtitle = seas[[5]], 
    title = NULL
  )
p9 <- show_avgseason(mods = toplo, doystr = 213, doyend = 338, yrstr = 2010, yrend = 2019, ylab = NULL) +
  coord_cartesian(ylim = ylim) + 
  labs(
    subtitle = seas[[6]],
    title = NULL
  )

>>>>

Examples of results provided by the trend tests are shown for station 34, with estimates of percent change between year pairs and seasonal trend results for January to June and July to December for the same year pairs (Figure \@ref(fig:trnddat)).  This figure illustrates the narrative descriptions that can be obtained for each test and the differences that can be observed depending on season.  Figures \@ref(fig:trnddat)a-c demonstrate that average chlorophyll at station 34 from 1990 to 2000 had a 268.4% increase, although the change is insignificant because of large uncertainties in the chlorophyll averages in the first year. Similarly, the change from 2000 to 2010 was an insignificant increase of 11%, followed by an insignificant decrease of 13.9% from 2010 to 2019.  Although these trends were insignificant, the results are based on a priori decisions of which years to compare.  Significant results may be observed for other year pairs.

@@

Spatial variation in trends suggested that most stations in South Bay had increasing chlorophyll from 1990 to 2010 following a decrease in recent years, although not all trends were significant and varied by location (Figure \@ref(fig:trndmap)a).  From 1990 to 2000, only station 18, 32, 34, and 36 had increasing trends that were not significant.  The magnitude of the chlorophyll increase in the southernmost stations (34, 36) was higher during this same period, although the trends were not significant.  From 2000 to 2010, significant increases were observed at the three northern-most stations (18, 21, and 22) and the three southern-most stations (32, 34, 36).  From 2010 to 2019, the three southern-most stations had significant decreases in annual chlorophyll, whereas station 24 had a significant increase during this period.  The spatial trends were further evaluated with mixed-meta regression analysis to provide additional context to the annual changes by evaluating trends from January to June and July to December (Figure \@ref(fig:trndmap)b).  For the January to June period, significant increases were observed at stations 32, 34, and 36 from 1991 to 2000 and station 18 from 2000 to 2010, whereas a significant decrease was observed at station 32 from 2010 to 2019. For the July to December period, significant increases were observed at stations 27 and 30 from 1991 to 2000 and stations 18, 21, 22, and 34 from 2000 to 2010, whereas significant decreases were observed at stations 32 and 34 from 2010 to 2019.

More discrete monthly periods in three month blocks were evaluated to explain a finer resolution of trends from the GAM results (Table \@ref(tab:seastrndtab)).  Many stations and time periods did not show any seasonal chlorophyll trends or only had one or two significant trends at this finer temporal scale.  The JAS (July, August, September) and OND (October, November, December) periods had significant trends that varied by decadal periods in the more northern stations (e.g., 18-27), whereas the southern stations tended to have more significant trends in the AMJ (April, May, June) and JAS periods. Accordingly, table \@ref(tab:seastrndtab) demonstrates the value of obtaining greater insight into seasonal changes by evaluating different blocks as compared to the larger seasonal blocks shown in Figure \@ref(fig:trndmap)b.

A final analysis evaluated a moving window comparison of seasonal trends to provide additional context of when significant changes are occurring at each station (Figure \@ref(fig:winchg)).  Results were obtained using a centered five year window to estimate the slope and significance of a seasonal trend of GAM results from January to June and July to December for each five-year period at each station (note that estimates at the far left and far right of each time series cannot be calculated).  As before, more significant trends were generally observed with the more southern stations, particularly in the July to December period.  Significant increases were observed for both January to June and July to December earlier in the period of record, whereas significant decreases were observed later in the record.  Most stations showed a shift around 2007 when trends switched from significantly increasing to significantly decreasing, particularly in the January to June period.

@@

Trends results from the mixed-meta regression method for each season and different time periods were compared to alternative trend analyses to demonstrate how different and potentially misleading conclusions about trends can arise from methods that insufficiently account for propagation of uncertainty.  As stated above, mixed-meta regression allow for full consideration of uncertainty in trend assessments by explicitly incorporating standard error of averages from results obtained from the GAMs and it is expected that more generalized methods that do not account for uncertainty may lead to different conclusions.  Figure \@ref(fig:trndcmp) shows estimates of chlorophyll change per year for the same seasonal periods and annual groupings as in figure \@ref(fig:trndmap)b.  In addition to mixed-meta analysis, two alternative and more general trend methods were used.  the first method shows slope estimates from a linear regression on annual averages of observed chlorophyll for the specified seasonal period and the second shows slope estimates from a simple linear regression model on annual averages derived from the GAM seasonal estimates.  The latter uses the same averages from the mixed-meta analysis, but is a simpler approach that does not account for the uncertainty in the average estimates.

The different trend analysis methods provided conflicting information on the magnitude and direction of the seasonal chlorophyll changes in each decade (Figure \@ref(fig:trndcmp)).  The slope estimates from the linear model applied to the observed data were understandably more variable than the slope estimate from the GAM average and mixed-meta methods, with much larger slopes observed especially at the more southern stations.  Slope estimates from the linear model applied to the averages from the GAMs as compared to the mixed-meta results were more identical, excluding some of the slope estimates for the southern stations.  The mixed-meta analyses had less significant trends compared to the average results for the same GAM estimates, which reflects the ability of the former to account for uncertainty in the average estimates when considering trends.  Importantly, completely opposite and/or varying significance of trends were observed between the methods, suggesting choice of analysis method can influence the certainty of conclusions.

@@

Explain increasing resolution of methods, e.g., test/base, to seasonal averages, to mixed-meta, to moving window approach

@@

p1 <- show_perchg(mods = toplo, baseyr = 1991, testyr = 2000, ylab = ylab) +
  labs(
    subtitle = paste0('(a) 1991: ', chg[[1, 2]],' ug/L, 2000: ', chg[[1, 3]], 'ug/L\nChange: ', chg[[1,4]], '%, ', chg[[1,5]]), 
    title = NULL
  )
p2 <- show_perchg(mods = toplo, baseyr = 2000, testyr = 2010, ylab = NULL) +
  labs(
    subtitle = paste0('(a) 2000: ', chg[[2, 2]],' ug/L, 2010: ', chg[[2, 3]], 'ug/L\nChange: ', chg[[2, 4]], '%, ', chg[[2, 5]]), 
    title = NULL
  )
p3 <- show_perchg(mods = toplo, baseyr = 2010, testyr = 2019, ylab = NULL) +
  labs(
    subtitle = paste0('(a) 2010: ', chg[[3, 2]],' ug/L, 2019: ', chg[[3, 3]], 'ug/L\nChange: ', chg[[3, 4]], '%, ', chg[[3, 5]]), 
    title = NULL
  )
p4 <- show_avgseason(mods = toplo, doystr = 41, doyend = 213, yrstr = 1991, yrend = 2000, ylab = ylab) +
  coord_cartesian(ylim = ylim) + 
  labs(
    subtitle = seas[[1]],
    title = NULL
  )
p5 <- show_avgseason(mods = toplo, doystr = 41, doyend = 213, yrstr = 2000, yrend = 2010, ylab = NULL) +
  coord_cartesian(ylim = ylim) + 
  labs(
    subtitle = seas[[2]], 
    title = NULL
  )
p6 <- show_avgseason(mods = toplo, doystr = 41, doyend = 213, yrstr = 2010, yrend = 2019, ylab = NULL) +
  coord_cartesian(ylim = ylim) +  
  labs(
    subtitle = seas[[3]], 
    title = NULL
  )
p7 <- show_avgseason(mods = toplo, doystr = 213, doyend = 338, yrstr = 1991, yrend = 2000, ylab = ylab) +
  coord_cartesian(ylim = ylim) + 
  labs(
    subtitle = seas[[4]], 
    title = NULL
  )
p8 <- show_avgseason(mods = toplo, doystr = 213, doyend = 338, yrstr = 2000, yrend = 2010, ylab = NULL) +
  coord_cartesian(ylim = ylim) + 
  labs(
    subtitle = seas[[5]], 
    title = NULL
  )
p9 <- show_avgseason(mods = toplo, doystr = 213, doyend = 338, yrstr = 2010, yrend = 2019, ylab = NULL) +
  coord_cartesian(ylim = ylim) + 
  labs(
    subtitle = seas[[6]],
    title = NULL
  )
